---
title: "Microbiome workshop: Data analysis and visualization"
author: "J. Gibbons"
output: html_document 
---
```{r setup}
  library(microbiome)
  library(ggplot2)
  library(ggpubr)
  library(knitr)
  source(here::here("Rsource/microbiome.day3.source.scripts.R"))

```
####Create a phyloseq object from DADA2 output####
```{r phyloseq-create}
infile_asv_counts<-here::here("Rdata/demo_asv_counts.tsv")
infile_asv_tax<-here::here("Rdata/demo_asvs_taxonomy.tsv")
infile_sample_data<-here::here("Rdata/made_up_sample_data.tsv")

df_asv_counts<-read.delim(infile_asv_counts)
df_asv_tax<-read.delim(infile_asv_tax)
df_sample_data<-read.delim(infile_sample_data)
#rownames(df_sample_data)<-df_sample_data[,1]
#Convert the data to matrix format
m_asv_counts<-as.matrix(df_asv_counts[2:length(df_asv_counts)])
rownames(m_asv_counts)<-df_asv_counts[,1]

m_asv_tax<-as.matrix(df_asv_tax[2:length(df_asv_tax)])
rownames(m_asv_tax)<-df_asv_tax[,1]

##There are 3 possible components to a phyloseq object: otu_table, sample_data, tax_table

pseq<-phyloseq(otu_table(m_asv_counts,taxa_are_rows = T),tax_table(m_asv_tax),sample_data(df_sample_data))


##Lets look at the data in our phyloseq object
##The functions used to create the components of a phyloseq object
##can be used to view the components of a phyloseq object
View(otu_table(pseq))
View(meta(pseq))
View(tax_table(pseq))
```

####Read in a dataset with interesting results and perform some initial processing####
```{r phyloseq-create}

infile<-here::here("Rdata/otu_table.biom")
infile_meta<-here::here("Rdata/meta_data.csv")


pseq<-read_biom2phyloseq(infile,metadata.file = infile_meta)

####Let's filter out some of the samples to simplify the analysis####
pseq<-subset_samples(pseq,group_name=="Without.comp.food" & diet !="Experimental infant formula")
####Summarize data at the Genus level. Not required, but convienent####
pseq<-aggregate_taxa(pseq,level="Genus")

```

##Inspect the phyloseq object
```{r inspect-phyloseq}
  View(otu_table(pseq))
  View(meta(pseq)) #Alternative method View(sample_data(pseq))
  View(tax_table(pseq))

```


####Calculate alpha diversity from milk study####
##Remember alpha diversity is calculated within a sample so normalization is not neccessary
```{r "alpha diversity"}
tab<-microbiome::diversity(pseq,index="all")
kable(tab)

##Let's create a dataframe for plotting the results
df_milk_meta<-meta(pseq)
##Add Shannon divesity data to the dataframe
df_milk_meta$Shannon<-tab$shannon

##Let's create a plot comparing the diversity between breast milk and formula
plot_div<-ggviolin(df_milk_meta,x="diet",y="Shannon",
             add="boxplot",fill="diet")
##Add stats (Wilcoxon test)
plot_div<-plot_div+stat_compare_means(comparisons = list(c("Breast milk","Standard infant formula")),
                          method="wilcox.test")
print(plot_div)

##Diversity is higher in the standard infant formula group
```

####Composition plot####
```{r composition}
##Compute the relative level for each taxa
pseq.rel<-microbiome::transform(pseq,"compositional")
##Remove rare OTUs
pseq.rel.core<-core(pseq.rel,detection = 0,prevalence = 0.5)
##aggregate taxa to the class level (so data will fit on graph)
pseq.rel.core.family<-aggregate_taxa(pseq.rel.core,level="Family")

##Recommend to save as 10X10
comp_plot<-plot_composition(pseq.rel.core.family,group_by="diet")+
  guides(fill=guide_legend(ncol=1))+
    labs(x = "Samples", y = "Relative abundance",
     title = "Relative abundance data")
print(comp_plot)
```


####Beta diversity and microbiome divergence####
```{r beta diversity}
##Result is sensitive to sample size. Recommended to subsample or bootstrap to
##avoid bias
##Rarefying data is inappropriate for differential abundance testing,
##but rarefying gives the best results when comparing sample similarity
##Calculate the divergences within the diet groups
pseq.rarefied<-rarefy_even_depth(pseq,rngseed=8675309,replace=F)
b.breast_milk<-divergence(subset_samples(pseq.rarefied,diet=="Breast milk"),method="bray")
b.formula<-divergence(subset_samples(pseq.rarefied,diet="Standard infant formula"))
boxplot(list(Breast_Milk=b.breast_milk,Formula=b.formula))
wilcox.test(x=b.breast_milk,y=b.formula,alternative = "less")
##The breast milk feed infants have more similar microbiomes to each other
##than the forumla fed infants to do each other
```


####Sample ordination####
```{r sample ordination}
library(vegan) #Perform statistics on ordination results
##Rarefying data is inappropriate for differential abundance testing,
##but rarefying gives the best results when comparing sample similarity
##Calculate the divergences within the diet groups
pseq.rarefied<-rarefy_even_depth(pseq,rngseed=8675309,replace=F)

##First lets plot the sample ordination
 ord_plot<-plot_landscape(pseq.rarefied,method="PCoA",distance = "bray",col="diet")+labs(title="PCoA/Bray-Curtis")
 print(ord_plot)
 
##We will use a PERMONVA to see if there are statistically distinguishable communties 
dist<-vegdist(t(otu_table(pseq.rarefied)),method="bray")

##Check if the variances are equal between the groups. If not assumtion for PERMOVNA not met
print(anova(betadisper(dist,meta(pseq.rarefied)$diet)))

##For this data the assumtion is not met, but here's the code
permonva_results<-adonis(t(otu_table(pseq.rarefied))~diet,data=meta(pseq.rarefied),
                 permutations = 100,method="bray")
print(permonva_results)

##Show coefficients for the top taxa separating the groups
permonva_result_coefs<-coefficients(permonva_results)
permonva_result_diet<-permonva_result_coefs["diet1",]

top.diet.coef<-permonva_result_diet[rev(order(abs(permonva_result_diet)))[1:5]]
par(mar=c(3,9,2,1)) #Setting margin sizes requires trial and error
barplot(sort(top.diet.coef),horiz=T,las=1,main="Top taxa")
```


####To use metagenomeSeq need to create a new data object called MRexperiment####
```{r Differential abundance analysis using metagenomeSeq}
library(metagenomeSeq)
pseq_mrexp<-phyloseq_to_metagenomeSeq(pseq)


###Keep samples with at least 1 count. Keep features with at least 30 count###
pseq_mrexp<-filterData(pseq_mrexp,present=30,depth=1)

###Normalization###
p<-cumNormStat(pseq_mrexp)
pseq_mrexp<-cumNorm(pseq_mrexp,p=p) #Calculate the scaling factors

###Create a design matrix and run statistical model###
phenotype_data<-pData(pseq_mrexp) #Get the metadata associated with this experiment

###Make breast milk the reference group (Reference is the denominator in logFC)###
phenotype_data$diet<-relevel(phenotype_data$diet,"Breast milk")
##A design matrix specicies what comparisons we want to make
##For more on design matrices check this youtube video: https://www.youtube.com/watch?v=CqLGvwi-5Pc
design_matrix<-model.matrix(~1+diet,data=phenotype_data) #Can drop 1, we're just explicitly showing an intercept

feature_fit_model<-fitFeatureModel(obj=pseq_mrexp,mod=design_matrix)

###Identify the differentially abundant microbes###
taxas<-fData(pseq_mrexp)$unique
sig_taxa<-MRcoefs(feature_fit_model,adjustMethod = "fdr",alpha = 0.1,taxa=taxas)
write.csv(sig_taxa,"sig_taxa_table_metagenomeSeq.csv")
####Graph the results###

##Identify the rows containing the abundance info for the significant taxa
sig_taxa_index<-which(rownames(MRcounts(pseq_mrexp)) %in% rownames(sig_taxa))
##Get the descriptions of the significant taxa
sig_taxa_desc<-rownames(MRcounts(pseq_mrexp))[sig_taxa_index]
##Find data locations for the 2 groups
classIndex<-list(Breast_milk=which(pData(pseq_mrexp)$diet=="Breast milk",arr.ind = T))
classIndex$Infant_formula<-which(pData(pseq_mrexp)=="Standard infant formula",arr.ind = T)


pdf("sig_taxa_metagenomeSeq.pdf")
for(i in seq(1,length(sig_taxa_index))){
  taxa_index<-sig_taxa_index[[i]]
  taxa_desc<-sig_taxa_desc[[i]]
  plotOTU(pseq_mrexp,otu=taxa_index,classIndex = classIndex,main=taxa_desc)
}

dev.off()

```

####Use DESeq2 to perform differential abundance analysis####
```{r differential abundance analysis using DESeq2}
library(DESeq2)
sample_data(pseq)$diet<-relevel(sample_data(pseq)$diet, "Breast milk")

##convert data to a format used by DESeq (group by diet)
pseq.deseq<-phyloseq_to_deseq2(pseq,~diet)

##Perform normalization and differential abundance test
pseq.deseq<-DESeq(pseq.deseq,test="Wald",fitType = "parametric")

##Get and filter the results
res<-results(pseq.deseq,cooksCutoff = F,pAdjustMethod = "BH",alpha=0.1)

res<-res[complete.cases(res),] ##Remove any NA results

##Write the results to file
res_table<-cbind(as(res,"data.frame"),as(tax_table(pseq)[rownames(res),],"matrix"))
write.csv(res_table,"sig_taxa_table_deseq2.csv")

##Graph the results

##Normalize the expersion data
pseq.clr<-microbiome::transform(pseq,"clr")
exp_data_clr<-abundances(pseq.clr)

##Keep only the differentially abundant microbes
exp_data_clr_sig<-exp_data_clr[which(rownames(exp_data_clr) %in% rownames(res)),]
exp_data_clr_sig<-melt(exp_data_clr_sig)
colnames(exp_data_clr_sig)<-c("Taxa","Sample","CLR")

##Extract the  metadata
df_meta<-meta(pseq.clr)
df_meta$Sample<-rownames(df_meta)
df_meta<-df_meta %>% select(Sample,diet)

df_dam<-left_join(exp_data_clr_sig,df_meta,by="Sample")

##Create the graphs
pdf("graph_deseq2_sig.pdf")
for(taxon in unique(df_dam$Taxa)){

  df_dam_taxon<-df_dam %>% filter(Taxa==taxon)
  plot<-ggplot(df_dam_taxon,aes(x=diet,y=CLR))+geom_dotplot(binaxis = "y",stackdir = "center")
  plot<-plot+ggtitle(taxon)+theme(plot.title = element_text(hjust=0.5))
  print(plot)
}
dev.off()


##Normalize the expersion data
pseq.norm<-microbiome::transform(pseq,"compositional")
exp_data_norm<-abundances(pseq.norm)

##Keep only the differentially abundant microbes
exp_data_norm_sig<-exp_data_norm[which(rownames(exp_data_norm) %in% rownames(res)),]
exp_data_norm_sig<-melt(exp_data_norm_sig)
colnames(exp_data_norm_sig)<-c("Taxa","Sample","Relative_Abundance")

##Extract the  metadata
df_meta<-meta(pseq.norm)
df_meta$Sample<-rownames(df_meta)
df_meta<-df_meta %>% select(Sample,diet)

df_dam<-left_join(exp_data_norm_sig,df_meta,by="Sample")

##Create the graphs
pdf("graph_deseq2_sig_ra.pdf")
for(taxon in unique(df_dam$Taxa)){
  
  df_dam_taxon<-df_dam %>% filter(Taxa==taxon)
  plot<-ggplot(df_dam_taxon,aes(x=diet,y=Relative_Abundance))+geom_dotplot(binaxis = "y",stackdir = "center")
  plot<-plot+ggtitle(taxon)+theme(plot.title = element_text(hjust=0.5))
  print(plot)
}
dev.off()
