---
title: "Basic_R"
author: "Chengqi(Charley) Wang"
output:
  html_document:
    df_print: paged
---
<br/>
*date: 1/29/2020*
###Arithmetic Operators
```
+            addition
-            subtraction
*            multiplication
/            division
^ or **	     exponentiation
x %% y	     modulus (x mod y) 5%%2 is 1
x %/% y	     integer division 5%/%2 is 2
```

###Logical Operators
```
<         less than
<=    	  less than or equal to
>     	  greater than
>=    	  greater than or equal to
==    	  exactly equal to
!=    	  not equal to
!x        Not x
x | y     x OR y
x & y     x AND y
isTRUE(x) test if X is TRUE
```
<br/><br/><br/>

###Data Types
R has a wide variety of data types including scalars, vectors,
matrices, data frames and lists.
<br/>

#####Creating new variables
```{r}
a <- 1
b <- 2
c <- a + b
```
<br/>

#####Vectors
```{r}
a <- c(1,2,5.3,6,-2,4) # numeric vector
b <- c("one","two","three") # character vector
c <- c(TRUE,TRUE,TRUE,FALSE,TRUE,FALSE) #logical vector
```
<br/>

####Refer to elements of a vector using subscripts
```{r}
a
a[c(2,4)] #2nd and 4th elements of vector
```
<br/>

####Matrix
All columns in a matrix must have the same mode(numeric, character, etc.) and the same length. The general format is

mymatrix <- matrix(vector, nrow=r, ncol=c, byrow=FALSE, 
   dimnames=list(char_vector_rownames, char_vector_colnames))

byrow=TRUE indicates that the matrix should be filled by rows. byrow=FALSE indicates that the matrix should be filled by columns (the default). dimnames provides optional labels for the columns and rows.

```{r}
# generates 5 x 4 numeric matrix 
y<-matrix(1:20, nrow=5,ncol=4)

# another example
cells <- c(1,26,24,68)
rnames <- c("R1", "R2")
cnames <- c("C1", "C2") 
mymatrix <- matrix(cells, nrow=2, ncol=2, byrow=TRUE,
  dimnames=list(rnames, cnames))

mymatrix
```
<br/>

####Identify rows, columns or elements using subscripts.
```{r}
x <- matrix(1:16,nrow = 4,ncol = 4)
x
x[,4] # 4th column of matrix
x[3,] # 3rd row of matrix 
x[2:4,1:3] # rows 2,3,4 of columns 1,2,3
```
<br/>

####Data Frames
A data frame is more general than a matrix, in that different columns can have different modes (numeric, character, factor, etc.). This is similar to SAS and SPSS datasets.
```{r}
expression <- runif(4) # random sampling four values from 0 to 1
gene <- c("gene1", "gene2", "gene3", "gene4")
f <- c(TRUE,TRUE,TRUE,FALSE)
mydata <- data.frame(Gene       = gene, 
                     Expression = expression, 
                     pick       =f)
print(mydata)
```
<br/>

####List
An ordered collection of objects (components). A list allows you to gather a variety of (possibly unrelated) objects under one name.
```{r}
# example of a list with 4 components - 
# a string, a numeric vector, a matrix, and a scaler 
w <- list(name = 'Marcus',  days = c('2ed', '4th', '20th', '28th'),
          exp = expression, height = '177cm')
w
```
Identify elements of a list using the [[]] convention.
```{r}
w[[3]] # 3rd component of the list
w[["exp"]] # component named exp in list
```
<br/>

####Factor
Tell R that a variable is nominal by making it a factor. The factor stores the nominal values as a vector of integers in the range [ 1... k ] (where k is the number of unique values in the nominal variable), and an internal vector of character strings (the original values) mapped to these integers.
```{r}
# variable gender with 20 "male" entries and 
# 30 "female" entries 
gender <- c(rep("male",20), rep("female", 30)) 
gender <- factor(gender) 
# stores gender as 20 1s and 30 2s and associates
# 1=female, 2=male internally (alphabetically)
# R now treats gender as a nominal variable 
summary(gender)
```
<br/>

####Useful Functions
```{r}
length(expression) # number of elements or components
str(expression)    # structure of an object 
class(expression)  # class or type of an object
names(expression)  # names

x1 <- 1:3; y1 <- 1:3
c(x1,y1)       # combine objects into a vector
cbind(x1,y1)   # combine objects as columns
rbind(x1,y1)   # combine objects as rows 

x1     # prints the object

ls()       # list current objects
rm(x1) # delete an object


```

